---
title: "R Notebook"
output: html_notebook
---
```{r}
library(tidyverse)
```
1.	Using the `read_csv()` function, read the GDP dataset into your notebook as a tibble called `gdp_df`. After reading it in, inspect the first 10 rows and then inspect the last 10 rows. 

```{r}
gdp_df <- read_csv('data/gdp_per_capita.csv')
```
```{r}
gdp_df |> 
  head(10)
```
```{r}
gdp_df |>
  tail(10)
```
2. Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.
```{r}
gdp_df <- gdp_df |>
  select(-'Value Footnotes') |>
  rename('Country' = 'Country or Area', 'GDP_Per_Capita' = 'Value')
gdp_df  
```
3. Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.
```{r}
gdp_df |> 
  pull(Year) |> 
  unique()
```
```{r}
  obs_per_year <- gdp_df |> 
                  group_by(Year) |> 
                  summarise(n())
obs_per_year
```

```{r}
obs_per_year <- obs_per_year |> 
                rename(count = 'n()')
 ggplot(obs_per_year, aes(x = Year, y = count)) + geom_col()
```
4. How many countries are represented in this dataset? Which countries have the fewest observations?
```{r}
 gdp_df |> 
  summarise(n_distinct(Country))
  
```
```{r}
obs_per_country <- gdp_df |> 
  group_by(Country) |> 
  summarise(n())  
  
obs_per_country |>
  rename(Count = 'n()') |> 
  arrange(Count)
```

